---
title: "Tracking code in Git/GitHub"
output: 
  html_document:
    highlight: espresso
    number_sections: yes
    theme: yeti
    toc: yes
    toc_depth: 2
    toc_float: yes
vignette: >
  %\VignetteIndexEntry{git}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE, out.width='70%', fig.show='hold'}
logo <- system.file("figures", "sph_cida_wm_blk.png", package="CIDAtools")
knitr::include_graphics(logo)
```

---

```{r setup, include=FALSE}
library(CIDAtools)
```

# Introduction

## Intro video

## What is Git?

## Notes

### Main vs Master

### Windows specific

# Getting Started

## Terminal / Git Bash

## Installing Git

## Creating GitHub Account

## Requesting CIDA GitHub Access

Following account creation, email your username to 
`person <at> cuanschutz <dot> edu`
to request access to CIDA's GitHub organization.

## Setting Up SSH

Copy/distill instructions here: 
<https://docs.github.com/en/authentication/connecting-to-github-with-ssh>

## Setting up SSO

Copy/distill instructions here: 
<https://docs.github.com/en/enterprise-cloud@latest/authentication/authenticating-with-saml-single-sign-on/about-authentication-with-saml-single-sign-on>

## Creating New Project

1. Create project folder on your local computer by calling 
`CIDAtools::CreateProject()` in `R`
2. Initialize git repository by navigating to folder (`cd Path/To/Folder`) then
using the command `git init`
3. Stage file changes `git add . `
3. Create initial commit with `git commit -m "Initial commit"`
4. Create empty repository on GitHub
    1. Instructions
    2. To do
    3. This
5. Add remote to local repository with 
`git remote -v git@github.com:CIDA-CSPH/<your-repository>.git`
6. Push work to GitHub with `git push origin main`

## Cloning Existing Project

1. Navigate to GitHub remote repository
2. Click green "Code &or;" button
3. Click "SSH" then copy link
4. Clone repository with `git clone git@github.com:CIDA-CSPH/<your-repository>.git`

# Regular Workflow

1. Add and commit changes
```{bash, eval=FALSE, echo=TRUE}
cd Path/To/Directory
git add .
git commit -m "My commit message"
```
2. Pull and merge any changes (see Handling Merge Conflicts below)
```{bash, eval=FALSE, echo=TRUE}
git pull origin main
## Fix any merge conflicts
git commit
```
3. Push all new commits
```{bash, eval=FALSE, echo=TRUE}
git push origin main
```

## Pushing updates

## Handling Merge conflicts

## Branching

# FAQ

# Git Kraken (Legacy)

---

```{r, echo=FALSE, out.width='70%', fig.show='hold'}
knitr::include_graphics(logo)
```
